Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-06-26 report/src/algebraic-structures.tex
Write about semi-closed structure

>>>>

\begin{remark}
  Hyland claims that `a $ \lambda $-theory is an algebraic theory $ L $ equipped with semi-closed structure'. By `semi-closed structure', he probably means the structure of a semi cartesian closed category on the Lawvere theory associated to $ L $ (Lemma \ref{lem:lawvere-clone}). Because a Lawvere theory has finite products, we even would have a weak cartesian closed category. For more information about weak cartesian closed categories, see Appendix \ref{ch:weak-cartesian-closed-categories}.

  In the appendix, we see that a $ \lambda $-theory structure on an algebraic theory $ L $ gives its associated Lawvere theory $ C $ a weak cartesian closed structure with the exponential object $ 1^1 $ equal to $ 1 $. Conversely, we can give $ L $ a $ \lambda $-theory structure from a weak cartesian closed structure on $ C $ where $ 1 $ is a reflexive object. Note that we really need $ 1 $ to be a reflexive object to make the construction work.

  However, note that a weak cartesian closed structure on $ C $ with $ 1 $ a reflexive object contains more information than a $ \lambda $-theory structure on $ L $: If we have a $ \lambda $-theory structure on $ L $, then construct from this a weak cartesian closed category structure with $ 1^1 = 1 $ on $ C $ and derive from this again a $ \lambda $-theory structure on $ L $, we get the same $ \lambda $-theory structure that we started out with. If, on the other hand, we go from a weak cartesian closed structure where $ 1 $ is a reflexive object to a $ \lambda $-theory structure and then to a weak cartesian closed structure again, we might end up with a different structure than the one that we started out with. This is because a weak cartesian closed structure is not necessarily unique up to isomorphism. In particular, we construct semi-exponential objects $ m^n $ which are equal to $ m $ for all $ m, n : C $, and derive their data, $ \mathrm{ev} $ and $ \mathrm{cur} $, from just the data of $ 1^1 $ in the original weak cartesian closed structure. Therefore, there are enough ways in which the resulting weak cartesian closed structure can differ from the original one.

  Note that here, we really need a choice for the exponential objects. It is not enough to just ask for the mere existence of semi-exponential objects. This is because we need the information contained in $ \mathrm{ev}: L_2 $ and $ \mathrm{cur} : L_{n + 1} \to L_n $ to define $ \lambda $ and $ \rho $.
\end{remark}

### 2024-06-26 report/src/weak-cartesian-closed.tex
Write about semi-closed structure

\chapter{Weak Cartesian Closed Categories}
In Definition \ref{def:endomorphism-theory}, we defined the endomorphism $ \lambda $-theory of a reflexive object in a cartesian closed category. However, as it turns out, we do not need all of the properties of a cartesian closed category. It suffices to consider a \textit{weak} cartesian closed category. In fact, even a `semi cartesian closed category' will suffice, but in this section, we will restrict ourselves to a weak cartesian closed category, as this notion suits our purposes well enough.

@@

\begin{definition}
  Let $ C $ a weak cartesian closed category. A \iindex{reflexive object} in $ C $ is an object $ U $ with morphisms $ f $ and $ g $, forming the following commutative diagram:
  \begin{center}
    \begin{tikzcd}
      U^U \arrow[dd, "\mathrm{cur}_{U^U}(\mathrm{ev})"'] \arrow[rd, "f"]\\
      & U \arrow[ld, "g"]\\
      U^U
    \end{tikzcd}
  \end{center}
\end{definition}

\begin{definition}
  If we have a reflexive object $ (U, f, g) $ in a weak cartesian closed category $ C $, we can endow the endomorphism algebraic theory $ E_C(U) $ (Definition \ref{def:endomorphism-theory}) with a $ \lambda $-theory structure with $ \beta $-equality, with
  \[ \rho(s) = ((s \cdot g) \times \id U) \cdot \mathrm{ev} : C(U^{n + 1}, U) \]
  and
  \[ \lambda(t) = \mathrm{cur}_{U^n}(t) \cdot f : C(U^n, U) \]
  for $ s: E(U)_n $ and $ t: E(U)_{n + 1} $.
\end{definition}

\begin{lemma}
  The definition above indeed yields a $ \lambda $-theory with $ \beta $-equality.
\end{lemma}
\begin{proof}
  We have
  \begin{align*}
    \rho(s \bullet t)
    &= ((\langle t_i \rangle_i \cdot (s \cdot g)) \times \id U) \cdot \mathrm{ev}\\
    &= (\langle t_i \rangle_i \times \id U) \cdot ((s \cdot g) \times \id U) \cdot \mathrm{ev}\\
    &= \rho(s) \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1, n + 1})),
  \end{align*}
  \begin{align*}
    \lambda(s) \bullet t
    &= \langle t_i \rangle_i \cdot \mathrm{cur}(t) \cdot f\\
    &= \mathrm{cur}_{U^n}((\langle t_i \rangle_i \times \id U) \cdot t) \cdot f\\
    &= \lambda(s \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1, n + 1})))
  \end{align*}
  and
  \begin{align*}
    \rho(\lambda(s))
    &= \langle \pi_1 \cdot \mathrm{cur}_{U^n}(t) \cdot \mathrm{cur}_{U^U}(\mathrm{ev}), \pi_2 \rangle \cdot \mathrm{ev}\\
    &= \langle \pi_1 \cdot \mathrm{cur}_{U^n}(t), \pi_2 \rangle \cdot \mathrm{ev}\\
    &= t.
  \end{align*}
\end{proof}

Now, we will show that the semi-exponential objects satisfy some properties that justify the use of the exponential notation. These properties allow us to construct semi-exponentials on products, from the semi-exponentials of their components.

@@

Now, we can show that giving an algebraic theory a $ \lambda $-theory structure amounts to endowing its corresponding Lawvere theory (Lemma \ref{lem:lawvere-clone}) with a weak cartesian closed structure.
\begin{lemma}
  Let $ L $ be an algebraic theory. A weak cartesian closed category structure on its associated Lawvere theory gives a $ \lambda $-theory structure on $ L $.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

\begin{lemma}
  Let $ L $ be a $ \lambda $-theory. Then its corresponding Lawvere theory $ C $ is a weak cartesian closed category.
\end{lemma}
\begin{proof}
  Note that by definition of a Lawvere theory, every $ n: C $ is the $ n $-fold product of $ 1 $. Therefore, binary products are given by addition on the natural numbers, and $ 0 $ is the terminal object.

  \TODO
\end{proof}

>>>>

\chapter{Weak Cartesian Closed Categories}\label{ch:weak-cartesian-closed-categories}
In Definition \ref{def:endomorphism-theory}, we defined the endomorphism $ \lambda $-theory of a reflexive object in a cartesian closed category. However, as it turns out, we do not need all of the properties of a cartesian closed category. It suffices to consider a \textit{weak} cartesian closed category. In fact, even a `semi cartesian closed category' will suffice, but in this section, we will restrict ourselves to a weak cartesian closed category, as this notion suits our purposes well enough.

@@

\begin{remark}
  Since the requirements for semi-exponential objects do not involve the usual universal properties, a weak cartesian closed structure (or a semi-exponential object) is not unique up to isomorphism. For example, see example 2.1 and 2.2 in \cite{hoofman-1993-semi-cartesian-closed} for two distinct weak cartesian closed structures on a category.
\end{remark}

\begin{definition}
  A \iindex{reflexive object} in a category $ C $ with binary products is an object $ U $ with an exponential $ U^U $ and morphisms $ f $ and $ g $, forming the following commutative diagram:
  \begin{center}
    \begin{tikzcd}
      U^U \arrow[dd, "\mathrm{cur}_{U^U}(\mathrm{ev})"'] \arrow[rd, "f"]\\
      & U \arrow[ld, "g"]\\
      U^U
    \end{tikzcd}
  \end{center}
\end{definition}

\begin{definition}
  If we have a reflexive object $ (U, f, g) $ in a category $ C $ with finite products, we can endow the endomorphism algebraic theory $ E_C(U) $ (Definition \ref{def:endomorphism-theory}) with a $ \lambda $-theory structure with $ \beta $-equality, with
  \[ \rho(s) = ((s \cdot g) \times \id U) \cdot \mathrm{ev} : C(U^{n + 1}, U) \]
  and
  \[ \lambda(t) = \mathrm{cur}_{U^n}(t) \cdot f : C(U^n, U) \]
  for $ s: E(U)_n $ and $ t: E(U)_{n + 1} $.
\end{definition}

\begin{lemma}
  The definition above indeed yields a $ \lambda $-theory with $ \beta $-equality.
\end{lemma}
\begin{proof}
  Note that for morphisms $ f: C(U^{m + 1}, U) $ and $ g: C(U^n, U)^m $,
  \[ (\langle g_i \rangle_i \times \id U) \cdot f = f \bullet (\iota_{n, 1}(g_1), \dots, \iota_{n, 1}(g_m), x_{n + 1}). \]
  Then
  \begin{align*}
    \rho(s \bullet t)
    &= ((\langle t_i \rangle_i \cdot s \cdot g) \times \id U) \cdot \mathrm{ev}\\
    &= (\langle t_i \rangle_i \times \id U) \cdot ((s \cdot g) \times \id U) \cdot \mathrm{ev}\\
    &= \rho(s) \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1})),
  \end{align*}
  \begin{align*}
    \lambda(s) \bullet t
    &= \langle t_i \rangle_i \cdot \mathrm{cur}(t) \cdot f\\
    &= \mathrm{cur}_{U^n}((\langle t_i \rangle_i \times \id U) \cdot t) \cdot f\\
    &= \lambda(s \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1}))),
  \end{align*}
  where we use the property of $ \mathrm{cur} $, and
  \begin{align*}
    \rho(\lambda(s))
    &= \langle \pi_1 \cdot \mathrm{cur}_{U^n}(t) \cdot \mathrm{cur}_{U^U}(\mathrm{ev}), \pi_2 \rangle \cdot \mathrm{ev}\\
    &= \langle \pi_1 \cdot \mathrm{cur}_{U^n}(t), \pi_2 \rangle \cdot \mathrm{ev}\\
    &= t
  \end{align*}
  where we use the property of $ \mathrm{ev} $ twice.
\end{proof}

We get the following as a special case, in which $ U^U $ is not only a retract of $ U $, but actually equal to $ U $.
\begin{corollary}
  Let $ L $ be an algebraic theory. Suppose that its associated Lawvere theory $ C $ is a weak cartesian closed category and suppose that for $ 1 : C $, we have $ 1^1 = 1 $. Then we can give $ L $ a $ \lambda $-theory structure with $ \beta $ equality, given by
  \[ \lambda_n(s) = \mathrm{cur}_n(s) \quad \text{and} \quad \rho(t) = \mathrm{ev} \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1})). \]
\end{corollary}

Now, we will show that the semi-exponential objects satisfy some properties that justify the use of the exponential notation. These properties allow us to construct semi-exponentials on products from the semi-exponentials of their components.
\begin{remark}
  Note that in a category $ C $ with a terminal object $ I $, we have exponential objects
  \[ A^I = A \quad \text{and} \quad I^A = I \]
  for all objects $ A $. Therefore, we get semi-exponential objects for free if we have a terminal object.
\end{remark}

@@

Now, we can show that giving an algebraic theory a $ \lambda $-theory structure gives its corresponding Lawvere theory (Lemma \ref{lem:lawvere-clone}) a weak cartesian closed structure.
\begin{lemma}
  Let $ L $ be a $ \lambda $-theory. Then its corresponding Lawvere theory $ C $ is a weak cartesian closed category.
\end{lemma}
\begin{proof}
  Note that by definition of a Lawvere theory, every $ n: C $ is the $ n $-fold product of $ 1 $. Therefore, binary products are given by addition on the natural numbers, and $ 0 $ is the terminal object.

  Now, we define $ 1^1 = 1 $, with
  \[ \mathrm{ev} = \rho(x_1) : C(2, 1) \quad \text{and} \quad \mathrm{cur}_n(f) = \lambda(f) : C(n, 1) \]
  for $ f: C(n + 1, 1) $. The following shows that this is indeed a semi-exponential object:
  \[
    \rho(x_1) \bullet (\lambda(f) \bullet g, h) = \rho(x_1 \bullet \lambda(f)) \bullet (g + (h)) = f \bullet (g + (h)) \]
  and
  \[ \lambda(f \bullet ((\iota_{n, 1}(g_i))_i + (x_{n + 1, n + 1}))) = \lambda(f) \bullet g \]
  Now, by the remark and the lemmas above, we have for $ m, n: C $ (both nonzero),
  \[ m^n = (1 \times \dots \times 1)^n = 1^n \times \dots \times 1^n \]
  and
  \[ 1^n = 1^{1 \times \dots \times 1} = ((1^1)^1\dots)^1 = 1 \]
  so $ C $ has semi-exponential objects and it is a weak cartesian closed category.
\end{proof}
